---
title: "2020-21 Annual Report"
output: html_notebook
---
This notebook is used to generate statistics and static graphs for the Annual Report. 
See the 2022-23_AnnualReport_InteractiveFigures.Rmd notebook for HTML figures.
---

Load packages
```{r message=FASLE}
library(tidyverse)
library(hablar)
library(rgdal)
library(sf)
library(zeallot)
library(ggalt)
library(maptools)
```

Import population data
```{r message=FALSE}
yearly_denominators <- readxl::read_excel("yearly_denominators_2022.xlsx") %>% 
  rename(year = year_of_death)
postcode_denominators <- read_csv("postcode_denominators_2022.csv")
postcode_denominators$adjusted_population <- as.numeric(postcode_denominators$adjusted_population)
```


Import data
```{r message=FALSE, warning=FALSE}
data = readxl::read_excel("2022-23_data.xlsx")
names(data) <- list("case_number","year","sex","region","SEIFA_disadvantage","cultural_background","remoteness","residential_status","state","cald","disability")
# convert fields to factors and reorder the age groups
data <- data %>% 
  convert(fct(case_number, sex, cultural_background, state, residential_status))
```

FIGURES:
1. Death rate by year of death and sex for all children and young people, South Australia 2005-2022
2. Death rate by metropolitan and inner rural regions for children and young people who were usually resident in South Australia, 2005-2022
3. Death rate by outer rural regions for children and young people who were usually resident in South Australia, 2005-2022
4. Death rate by index of relative socioeconomic disadvantage for all children and young people who died in South Australia, 2005-2022
5. Death rate by cultural background for all children and young people, South Australia 2005 - 2022


Analyses and figures by section
---
1. Child Deaths South Australia 2005–2021
  1.2.	Rates and patterns of death
```{r}
# calculate death rates by sex
rates <- data %>%
  #filter(sex %in% c("Male","Female")) %>% 
  count(year, sex) %>% # Numerator
  mutate(type = 2) %>% 
  bind_rows(data %>%
              count(year) %>%
              mutate(sex = "Total",
                     type = 1)) %>% # Total numerator
  left_join(yearly_denominators %>% # join to the population data
              group_by(year, sex) %>%
              summarise_at(c("adjusted_population"), sum) %>%
              ungroup() %>%
              bind_rows(yearly_denominators %>%
                          group_by(year) %>%
                          summarise_at(c("adjusted_population"), sum) %>%
                          ungroup() %>%
                          mutate(sex = "Total")), # Total denominator
            by = c("year", "sex")) %>% 
  mutate(rate = (n/adjusted_population)*100000, # calculate the rates
         type = as.factor(type)) %>%
  rename("Death rate per 100,000 resident population" = rate)

# Data table 1
write_csv(rates %>%
            rename(Deaths = n, Year = year, Sex = sex,
                   "Estimated resident population" = adjusted_population) %>%
            select(-c(type)),
          "Tables/Data_table_1.csv")

# generate the sentence for the report
paste("In 2022, the number of deaths of children and young people was", filter(rates, year == 2022 & sex == "Total")$n)
```
  
```{r}
paste("During this 18 year period, the average yearly population of children and young people aged 0 to 17 was", 
      round(mean(filter(rates, sex == "Total")$adjusted_population),0))
```

```{r}
data %>% group_by(year) %>% count()
```

  Figure 1
```{r}
custom_colours <- c("#000000", "#FEB627", "#7E7E7E", "#27B9FE", "#FFFFFF")
# generate the static figure for the PDF report. this is performed for all the figures - the interactive ones are generated in the other notebook
(fig_one <- ggplot(data = rates, 
                   aes(x = year, y = `Death rate per 100,000 resident population`, 
                       colour = sex, linetype = type)) + 
  geom_line(size = 2.5) + 
  scale_colour_manual(values = custom_colours[c(2, 4, 3)]) + 
  theme_minimal() + 
  guides(linetype = "none") + 
  labs(y = "Death rate per 100,000 resident population", x = "Year") + 
  theme(axis.title.y = element_text(margin = margin(t = 0, r = 20, b = 0, l = 0)),
        legend.position = c(0.793,0.878),
        legend.title = element_blank()) +
  coord_cartesian(ylim = c(19, 45)) + 
  scale_x_continuous(breaks = seq(2006, 2022, by = 2),
                     minor_breaks = seq(2005, 2022, 1)) +
  scale_y_continuous(breaks = seq(20, 50, by = 5)) + 
  theme(text = element_text(size = 14)) 
)

# save the figure in both tiff and png format
ggsave(plot = fig_one,
       path = "Figures",
       filename = "Figure 1 Death rate by year of death and sex for all children.tiff",
       width = 19.5,
       height = 13.5,
       units = "cm",
       device = "tiff",
       dpi = 320)

ggsave(plot = fig_one,
       path = "Figures",
       filename = "figure1.png",
       width = 19.5,
       height = 13.5,
       units = "cm",
       device = "png",
       dpi = 320)
```

 
  1.2.1. Death rates by region
```{r message=FALSE, echo=FALSE}
# Deaths rates by region - load and transform data
# Read in shapefiles
#shp_df <- st_read("gov_regions\\SAGovtRegions_GDA94.shp") # Gov regions shapefile
shp_df <- readOGR(dsn = "gov_regions", layer = "SAGovtRegions_GDA2020")

# add column of row names to data object
shp_df@data$id <- rownames(shp_df@data)
# create dataframe from spatial object
regions_points <- fortify(shp_df, region = "id")
# merge the fortified data with the spatial data
regions_df <- merge(regions_points, shp_df@data, by = "id")

aus_shp <- readOGR(dsn = "shapes", layer = "aust_cd66states") # Australia shapefile

sa_shp <- subset(aus_shp, STE == 4) # Extract SA shape

numerator_region <- data %>%
  filter(!is.na(region)) %>% 
  count(region)

denominator_region <- postcode_denominators %>%
  filter(!is.na(region)) %>%
  group_by(region) %>%
  summarize(adjusted_population = sum(adjusted_population, na.rm = TRUE)) %>%
  ungroup()

rates_regions <- numerator_region %>%
  left_join(denominator_region, by = "region") %>%
  mutate(rate = (n/adjusted_population)*100000) 

#Data table 2
write_csv(rates_regions %>%
            mutate(adjusted_population = adjusted_population/18) %>% 
            rename("Number of deaths" = n,
                   "Estimated resident population (mean)" = adjusted_population,
                   "Death rate per 100,000 resident population" = rate,
                   "SA Government Region" = region),
          "Tables/Data_table_2.csv")

map_df <- rates_regions %>% 
  left_join(regions_df, by = "region") %>%
  rename("Rate per 100,000\nresident population" = rate)

# set the coordinates for the labels
centre_names <- aggregate(data = map_df, cbind(long, lat) ~ region, FUN = mean) 
centre_names[8, 2] <- 139.7413
centre_names[8, 3] <- -34.8774
centre_names[6, 2] <- 137.2142
centre_names[6, 3] <- -35.7752
centre_names[3, 2] <- 138.79
centre_names[3, 3] <- -34.9
centre_names[11, 2] <- 138.4
```

Plot death rates by region - Metropolitan
```{r}
centre_names[11, 3] <- -34.85
centre_names[2, 3] <- -34.5
centre_names[10, 2] <- 138.465

(fig_two <- ggplot() + 
  geom_polygon(data = filter(map_df, region %in% c("Barossa, Light and Lower North", "Northern Adelaide", "Eastern Adelaide", 
                                                   "Western Adelaide", "Adelaide Hills", "Southern Adelaide")), 
               aes(x = long, y = lat, group = region, fill = `Rate per 100,000\nresident population`), colour = "white") +
   scale_fill_gradient(limits= c(17, 35), low = custom_colours[5], high = custom_colours[2]) + 
  geom_text(data = filter(centre_names, region %in% c("Barossa, Light and Lower North", "Northern Adelaide", 
                                                  "Eastern Adelaide", "Western Adelaide", "Adelaide Hills", "Southern Adelaide")), 
            aes(x = long, y = lat, label = region), size = 5) + theme_void() + 
  geom_polygon(data = sa_shp, aes(x = long, y = lat, group = group), colour = "black", fill = NA) + 
  coord_map(xlim = c(138, 139.2), ylim = c(-35.4, -34.1)) +
  theme(text = element_text(size = 14))
)

ggsave(plot = fig_two,
       path = "Figures",
       "Figure 2 Death rate by region of metropolitan South Australia for children who were usual residents.tiff",
       width = 19.5,
       height = 16,
       units = "cm",
       device = "tiff",
       dpi = 320)

ggsave(plot = fig_two,
       path = "Figures",
       "figure2.png",
       width = 19.5,
       height = 16,
       units = "cm",
       device = "png",
       dpi = 320)
```

Plot death rates by region - Non-metropolitan
```{r}
(fig_three <- ggplot() + 
  geom_polygon(data = filter(map_df, !region %in% c("Barossa, Light and Lower North", "Northern Adelaide", "Eastern Adelaide", 
                                                   "Western Adelaide", "Adelaide Hills", "Southern Adelaide")), 
               aes(x = long, y = lat, group = region, fill = `Rate per 100,000\nresident population`), colour = "white") +
   scale_fill_gradient(limits= c(25, 95), low = custom_colours[5], high = custom_colours[2]) + 
  geom_text(data = filter(centre_names, !region %in% c("Barossa, Light and Lower North", "Northern Adelaide", 
                                                  "Eastern Adelaide", "Western Adelaide", "Adelaide Hills", "Southern Adelaide")), 
            aes(x = long, y = lat, label = region), size = 5) + theme_void() + 
  geom_polygon(data = sa_shp, aes(x = long, y = lat, group = group), colour = "black", fill = NA) + 
  coord_map(xlim = c(128, 142), ylim = c(-25.2, -39)) +
  theme(text = element_text(size = 14))
)

ggsave(plot = fig_three,
       path = "Figures",
       "Figure 3 Death rate by region of non-metropolitan South Australia for children who were usual residents.tiff",
       width = 19.5,
       height = 16,
       units = "cm",
       device = "tiff",
       dpi = 320)

ggsave(plot = fig_three,
       path = "Figures",
       "figure3.png",
       width = 19.5,
       height = 16,
       units = "cm",
       device = "png",
       dpi = 320)
```


  1.2.3.	Death rates and socioeconomic disadvantage
Death rate by SEIFA quintile - disadvantage
```{r}
numerator_disadvantage <- data %>%
  filter(!is.na(SEIFA_disadvantage)) %>% 
  count(year, SEIFA_disadvantage)

denominator_disadvantage <- postcode_denominators %>%
  group_by(year, SEIFA_disadvantage) %>%
  summarise_at(c("adjusted_population"), sum) %>%
  ungroup()

# calculate the rates
rates_disadvantage <- numerator_disadvantage %>%
  left_join(denominator_disadvantage, by = c("year", "SEIFA_disadvantage")) %>%
  mutate(rate = (n/adjusted_population)*100000) %>%
  group_by(SEIFA_disadvantage) %>%
  summarise(rate = mean(rate), Deaths = sum(n)) %>%
  rename("Death rate per 100,000 resident population" = rate, "Relative Socio-Economic Disadvantage Quintile" = SEIFA_disadvantage)

#Data table 3
write_csv(rates_disadvantage, 
          "Tables/Data_table_3.csv")
```

```{r}
cbPalette <- c("#999999", "#56B4E9", "#E69F00", "#0072B2", "#CC79A7")
seifa_colours <- c("#FEB627", "#27B9FE", "mediumseagreen", "sienna2", "slateblue2")

(fig_four <- ggplot(data = rates_disadvantage, 
                   aes(x = as.factor(`Relative Socio-Economic Disadvantage Quintile`), 
                       y = `Death rate per 100,000 resident population`,
                       fill = as.factor(`Relative Socio-Economic Disadvantage Quintile`))) + 
  geom_bar(stat = "identity", show.legend = FALSE) +
  labs(x = "Index of Relative Socio-Economic Disadvantage") +
  scale_fill_manual(values = seifa_colours) +
  coord_cartesian(ylim = c(0, 41)) + 
  scale_y_continuous(breaks = seq(0, 50, by = 10)) +
  scale_x_discrete(labels=c("1\n(most disadvantaged)", 2, 3, 4, "5\n(least disadvantaged)")) +
  theme_minimal() + 
  theme(text = element_text(size = 14),
        axis.title.x = element_text(vjust=-0.5))
)

ggsave(plot = fig_four,
       path = "Figures",
       filename = "Figure 4 Death rate by Relative Socio-Economic Disadvantage Quintile.tiff",
       width = 19.5,
       height = 13.5,
       units = "cm",
       device = "tiff",
       dpi = 320)

ggsave(plot = fig_four,
       path = "Figures",
       filename = "figure4.png",
       width = 19.5,
       height = 13.5,
       units = "cm",
       device = "png",
       dpi = 320)
```


1.5.	Deaths of Aboriginal children and young people
```{r}
# calculate death rates for Aboriginal and non-Aboriginal children and young people
rates_atsi <- data %>%
  count(year, cultural_background) %>% # Numerator
  bind_rows(data %>%
            count(year) %>%
            mutate(cultural_background = "Total")) %>% # Total numerator
  left_join(yearly_denominators %>% # join the population data
            mutate(cultural_background = fct_recode(cultural_background,"Aboriginal or Torres Strait Islander"="ATSI")) %>% 
            group_by(year, cultural_background) %>%
            summarise_at(c("adjusted_population"), sum) %>%
            ungroup() %>%
            bind_rows(yearly_denominators %>%
                      group_by(year) %>%
                      summarise_at(c("adjusted_population"), sum) %>%
                      ungroup() %>%
                      mutate(cultural_background = "Total")), # Total denominator
            by = c("year", "cultural_background")) %>% 
  mutate(rate = (n/adjusted_population)*100000,
         cultural_background = fct_recode(cultural_background, Aboriginal = "Aboriginal or Torres Strait Islander")) %>%
  rename("Death rate per 100,000 resident population" = rate)

#Data table 4
write_csv(rates_atsi %>% 
            rename("Year" = year,
                   "Cultural Background" = cultural_background,
                   "Number of deaths" = n,
                   "Estimated resident population" = adjusted_population),
          "Tables/Data_table_4.csv")

# calculate the average population (yearly) of children and young people over the reporting period
average_populations <- yearly_denominators %>% 
  group_by(year, cultural_background) %>%
  summarise(n = sum(adjusted_population)) %>% 
  group_by(cultural_background) %>% 
  summarise(population = mean(n))

# and the average death rate
average_death_rates <- rates_atsi %>%
  filter(cultural_background != "Total") %>% 
  group_by(cultural_background) %>% 
  summarise(mean(`Death rate per 100,000 resident population`))
```

```{r}
paste0("During the period 2005 to 2022, Aboriginal children and young people constituted only ",
       round(average_populations[1,"population"]/sum(average_populations$population)*100,1), "% ", 
       "of the South Australian population of children and young people, but they accounted for ", 
       round(nrow(filter(data, cultural_background == "Aboriginal or Torres Strait Islander"))/nrow(data)*100,1), "% of child deaths.")
```
```{r}
# do the same as above, but only including children who were normally resident in SA
rates_atsi_residents <- data %>%
  filter(state == "SA") %>% 
  count(year, cultural_background) %>% # Numerator
  bind_rows(data %>%
            count(year) %>%
            mutate(cultural_background = "Total")) %>% # Total numerator
  left_join(yearly_denominators %>%
            mutate(cultural_background = fct_recode(cultural_background,"Aboriginal or Torres Strait Islander"="ATSI")) %>% 
            group_by(year, cultural_background) %>%
            summarise_at(c("adjusted_population"), sum) %>%
            ungroup() %>%
            bind_rows(yearly_denominators %>% # Denominator
                      group_by(year) %>%
                      summarise_at(c("adjusted_population"), sum) %>%
                      ungroup() %>%
                      mutate(cultural_background = "Total")), # Total denominator
            by = c("year", "cultural_background")) %>% 
  mutate(rate = (n/adjusted_population)*100000) %>% 
  rename("Death rate per 100,000 resident population" = rate)

average_death_rates_residents <- rates_atsi_residents %>%
  filter(cultural_background != "Total") %>% 
  group_by(cultural_background) %>% 
  summarise(mean(`Death rate per 100,000 resident population`))

paste0("The rate of death for all Aboriginal children and young people who died in South Australia was ", 
       round(average_death_rates[1,2],1), 
       " deaths per 100,000. For Aboriginal children and young people who were usually resident in South Australia, the death rate was ", 
       round(average_death_rates_residents[1,2],1),
       " deaths per 100,000 over the same period.")
```

```{r}
paste0("The rate of death for non-Aboriginal children and young people was ",
       round(average_death_rates[2,2],1), 
       " deaths per 100,000. The rate of death for non-Aboriginal children and young people usually resident in South Australia was ",
       round(average_death_rates_residents[2,2],1), " deaths per 100,000.")
```

Figure 5
```{r}
(fig_five <- ggplot(filter(rates_atsi, cultural_background %in% c("Aboriginal", "Other"))) + 
  geom_line(aes(x = year, 
                y = `Death rate per 100,000 resident population`, 
                color = cultural_background), 
            size=1.5) +
  xlab("Year") +
  scale_color_manual(values=c("#FEB627", "#27B9FE")) + 
  scale_x_continuous(breaks = seq(2006, 2022, by = 2),
                     minor_breaks = seq(2005, 2022, 1)) +
  theme_minimal() + 
  theme(text = element_text(size = 14),
        legend.title = element_blank(),
        legend.position = c(0.81, 0.93)) 
)

ggsave(plot = fig_five,
       path = "Figures",
       filename = "Figure 5 Death rate by cultural background, South Australia 2005-2021.tiff",
       width = 19.5,
       height = 13.5,
       units = "cm",
       device = "tiff",
       dpi = 320)

ggsave(plot = fig_five,
       path = "Figures",
       filename = "figure5.png",
       width = 19.5,
       height = 13.5,
       units = "cm",
       device = "png",
       dpi = 320)
```
